---
title: "Cliniface_vs_Meshmonk"
author: "Abiodun Olowo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)      # For file writing functions
library(ggplot2)    # For plotting
library(dplyr)      # For data manipulation
```
### Additional Debugging Tips:


# Chunk 1A: Load and Validate Mesh Files

```{r Load Vertices, echo=FALSE, message=FALSE, warning=FALSE}
# === Load required libraries ===
library(fs)
library(purrr)
library(stringr)
library(tibble)
library(dplyr)
library(geomorph)

# === Define folders ===
folder_MM <- "C:/Users/ayo5146/Documents/CLINIFACE_VS_MESHMONK/Meshmonk"
folder_CF <- "C:/Users/ayo5146/Documents/CLINIFACE_VS_MESHMONK/Cliniface"
output_dir <- "C:/Users/ayo5146/Documents/CLINIFACE_VS_MESHMONK"
dir.create(output_dir, showWarnings = FALSE)

# === List OBJ files ===
files_MM <- dir_ls(folder_MM, regexp = "\\.obj$")
files_CF <- dir_ls(folder_CF, regexp = "\\.obj$")

# === Extract base names ===
get_basename <- function(x) tools::file_path_sans_ext(basename(x))
basenames_MM <- map_chr(files_MM, get_basename)
basenames_CF <- map_chr(files_CF, get_basename)

# === Validate filenames ===
valid_pattern <- "^[FM][EKIYS][0-9]{3}(MM|CF)$"  # Updated regex
valid_pops <- c("E", "K", "I", "Y", "S")        # Valid population codes

# Validate format and population codes
validate_name <- function(names, method) {
  format_ok <- str_detect(names, valid_pattern)
  pop_ok <- str_sub(names, 2, 2) %in% valid_pops
  format_ok & pop_ok
}

valid_MM <- validate_name(basenames_MM, "MM")
valid_CF <- validate_name(basenames_CF, "CF")

valid_files_MM <- files_MM[valid_MM]
valid_files_CF <- files_CF[valid_CF]
invalid_files_MM <- files_MM[!valid_MM]
invalid_files_CF <- files_CF[!valid_CF]

# === Report and export invalids ===
write_lines(invalid_files_MM, file.path(output_dir, "invalid_filenames_MM.txt"))
write_lines(invalid_files_CF, file.path(output_dir, "invalid_filenames_CF.txt"))
cat("✅ MM valid:", length(valid_files_MM), " | invalid:", length(invalid_files_MM), "\n")
cat("✅ CF valid:", length(valid_files_CF), " | invalid:", length(invalid_files_CF), "\n")

# === Extract metadata ===
extract_metadata <- function(names) {
  tibble(
    filename = names,
    Sex = str_sub(names, 1, 1),
    Pop = str_sub(names, 2, 2),
    Method = str_sub(names, 6, 7),
    BaseID = str_sub(names, 1, 5)  # Sex + Pop + 3-digit ID
  )
}

metadata_MM <- extract_metadata(basenames_MM[valid_MM])
metadata_CF <- extract_metadata(basenames_CF[valid_CF])
metadata <- bind_rows(metadata_MM, metadata_CF)

# === Load vertex data ===
source("load_vertices_unfilteredc.R")
meshes_MM <- map(valid_files_MM, load_all_vertices_unfiltered)
meshes_CF <- map(valid_files_CF, load_all_vertices_unfiltered)

# === Combine ===
all_meshes_raw <- c(meshes_MM, meshes_CF)
final_valid_files <- c(valid_files_MM, valid_files_CF)

# === Check for NA and consistent vertex count ===
has_na <- map_lgl(all_meshes_raw, ~ any(is.na(.x)))
all_meshes_raw <- all_meshes_raw[!has_na]
final_valid_files <- final_valid_files[!has_na]

vertex_counts <- map_int(all_meshes_raw, nrow)
if (length(unique(vertex_counts)) > 1) {
  stop("❌ Meshes have inconsistent vertex counts.")
}
n_vertices <- vertex_counts[1]
n_shapes <- length(all_meshes_raw)
```
# Chunk 1B: Build Metadata and Run GPA
```{r Build Metadata and GPA, echo=FALSE, message=FALSE, warning=FALSE}
# === Build shape array ===
vertex_array <- array(NA, dim = c(n_vertices, 3, n_shapes))
for (i in seq_len(n_shapes)) {
  vertex_array[,,i] <- all_meshes_raw[[i]]
}

# === Build metadata (Updated for new naming convention) ===
extract_info <- function(name) {
  tibble(
    filename = name,
    Sex = str_sub(name, 1, 1),        # First character: F/M
    Pop = str_sub(name, 2, 2),        # Second character: E/K/I/Y/S
    Method = str_sub(name, 6, 7),      # Last 2 characters: MM/CF
    BaseID = str_sub(name, 1, 5)      # First 5 chars (e.g., FE001)
  )
}

# Use final_valid_files (from Chunk 1A) to ensure consistency
metadata <- map_dfr(map_chr(final_valid_files, get_basename), extract_info)

# === Validate metadata ===
# Check for valid population codes
valid_pops <- c("E", "K", "I", "Y", "S")
invalid_pops <- metadata %>% 
  filter(!Pop %in% valid_pops) %>% 
  distinct(Pop)

if (nrow(invalid_pops) > 0) {
  warning("⚠️ Invalid population codes detected: ", paste(invalid_pops$Pop, collapse = ", "))
}

# === Final sanity checks ===
stopifnot(
  nrow(metadata) == dim(vertex_array)[3],
  all(metadata$Method %in% c("MM", "CF")),
  all(metadata$Sex %in% c("F", "M"))
)
cat("✅ Loaded and aligned:", nrow(metadata), "meshes\n")

# === Run GPA ===
cat("🔄 Performing Generalized Procrustes Alignment...\n")
gpa_result <- gpagen(vertex_array, print.progress = TRUE)
aligned_coords <- gpa_result$coords
consensus_shape <- mshape(aligned_coords)
cat("✅ GPA completed.\n")
# Quick verification
glimpse(metadata)
table(metadata$Pop, metadata$Method)
stopifnot(dim(aligned_coords)[3] == nrow(metadata))
```

# === Export key objects for Chunk 2 ===
# (aligned_coords and metadata are now ready for pairwise comparisons)

# === Chunk 2: Find Matched Pairs and Compute Distances ===

# Chunk 2: Corrected Pairwise Comparisons
```{r Get base IDs, echo=FALSE, message=FALSE, warning=FALSE}
# === Step 1: Verify Metadata Structure ===
required_cols <- c("BaseID", "Method", "filename")
missing_cols <- setdiff(required_cols, colnames(metadata))
if (length(missing_cols) > 0) {
  stop("❌ Missing columns in metadata: ", paste(missing_cols, collapse = ", "))
}

# === Step 2: Reorganize Metadata ===
# Sort metadata to ensure MM and CF scans are adjacent for each BaseID
metadata_sorted <- metadata %>%
  arrange(BaseID, Method)  # MM comes before CF in alphabetical order

# === Step 3: Find Valid MM-CF Pairs ===
pairwise_comparisons <- metadata_sorted %>%
  # Add original row numbers after sorting
  mutate(row_num = row_number()) %>%  
  group_by(BaseID) %>%
  # Ensure exactly 1 MM and 1 CF scan per subject
  filter(n() == 2, all(c("MM", "CF") %in% Method)) %>%
  # Extract row numbers - MM is always first in each group due to sorting
  summarise(
    row_MM = row_num[Method == "MM"],  # Row of MM scan
    row_CF = row_num[Method == "CF"],  # Row of CF scan
    .groups = "drop"
  )

# === Check 1: Confirm Pairs Exist ===
if (nrow(pairwise_comparisons) == 0) {
  stop("❌ No MM-CF pairs found. Check metadata for inconsistent BaseIDs.")
} else {
  cat("✅ Found", nrow(pairwise_comparisons), "valid MM-CF pairs.\n")
}

# === Check 2: Validate Row Numbers ===
max_row <- max(c(pairwise_comparisons$row_MM, pairwise_comparisons$row_CF))
if (max_row > dim(aligned_coords)[3]) {
  stop(
    "❌ Row numbers exceed aligned_coords dimensions.\n",
    "Max row:", max_row, "| aligned_coords size:", dim(aligned_coords)[3]
  )
}

# === Step 4: Compute Distances ===
pairwise_comparisons <- pairwise_comparisons %>%
  mutate(
    # Calculate RMSE (vertex-wise Euclidean distances)
    RMSE = map2_dbl(
      row_MM, row_CF, 
      ~ sqrt(mean((aligned_coords[, , .x] - aligned_coords[, , .y])^2))
    ),
    # Calculate Procrustes distance (global shape difference)
    ProcrustesDistance = map2_dbl(
      row_MM, row_CF,
      ~ sqrt(sum((aligned_coords[, , .x] - aligned_coords[, , .y])^2))
    )
  )

# === Check 3: Verify Distance Variation ===
if (sd(pairwise_comparisons$RMSE) < 0.0001) {
  warning("⚠️ RMSE values show suspiciously low variation. Verify data pairing.")
}

# === Final Output ===
glimpse(pairwise_comparisons)
cat("✅ Chunk 2 completed successfully.\n")

# Export results
write_csv(pairwise_comparisons, file.path(output_dir, "pairwise_distances.csv"))
```


# Chunk 3: Statistical Analysis Preparation
```{r Stats_MM_CF, echo=FALSE, message=FALSE, warning=FALSE}
# === Data Validation Checks ===
# 1. Verify metadata structure
required_meta_cols <- c("BaseID", "Method", "Sex", "Pop")
missing_meta_cols <- setdiff(required_meta_cols, colnames(metadata))
if (length(missing_meta_cols) > 0) {
  stop("❌ Missing columns in metadata: ", paste(missing_meta_cols, collapse = ", "))
}

# 2. Check aligned coordinates dimensions
cat("Dimensions of aligned_coords:", dim(aligned_coords), "\n")  # Should be [n_vertices, 3, n_shapes]
stopifnot(
  "aligned_coords should be 3D array" = length(dim(aligned_coords)) == 3,
  "Metadata/aligned_coords mismatch" = nrow(metadata) == dim(aligned_coords)[3]
)

# 3. Validate pairwise comparisons structure
cat("\n=== Pairwise Comparisons Structure ===\n")
glimpse(pairwise_comparisons)
stopifnot(
  all(c("BaseID", "row_MM", "row_CF", "RMSE", "ProcrustesDistance") %in% 
        names(pairwise_comparisons))
)

# === Data Quality Checks ===
# 1. Verify RMSE calculation
idx1 <- pairwise_comparisons$row_MM[1]
idx2 <- pairwise_comparisons$row_CF[1]
test_RMSE <- sqrt(mean((aligned_coords[, , idx1] - aligned_coords[, , idx2])^2))
cat("\nTest RMSE for first pair:", test_RMSE, "| Stored value:", 
    pairwise_comparisons$RMSE[1], "\n")
stopifnot(abs(test_RMSE - pairwise_comparisons$RMSE[1]) < 1e-10)

# 2. Check vertex count consistency
vertex_counts <- map_int(1:dim(aligned_coords)[3], ~ nrow(aligned_coords[, , .x]))
if (length(unique(vertex_counts)) != 1) {
  warning("⚠️ Inconsistent vertex counts detected: ", 
          paste(range(vertex_counts), collapse = "-"))
} else {
  cat("✔ Consistent vertex count:", vertex_counts[1], "\n")
}

# === Merge Metadata ===
pairwise_enhanced <- pairwise_comparisons %>%
  left_join(
    metadata %>% 
      filter(Method == "MM") %>%  # Only need one record per BaseID
      select(BaseID, Sex, Pop),
    by = "BaseID"
  )

# Verify merge success
if (any(is.na(pairwise_enhanced$Sex))) {
  warning("⚠️ Missing demographic data for some subjects")
}

# === Prepare Final Results ===
results_columns <- c("BaseID", "ProcrustesDistance", "RMSE", "Sex", "Pop")
final_results <- pairwise_enhanced %>%
  select(all_of(results_columns)) %>%
  rename(
    ProcrustesDistance_MM_vs_CF = ProcrustesDistance,
    RMSE_MM_vs_CF = RMSE
  )

# === Output Preview ===
cat("\n=== Final Results Structure ===\n")
glimpse(final_results)
cat("\n✅ Chunk 3 completed successfully. Ready for statistical analysis.\n")

# Export prepared data
write_csv(final_results, file.path(output_dir, "analysis_ready_data.csv"))
```


# Chunk 4: Publication-ready Outputs
```{r Publication-ready Outputs, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gt)
library(ggridges)
library(kableExtra)

# === Data Validation ===
stopifnot(
  exists("final_results"),  # From Chunk 3
  exists("aligned_coords"),
  file.exists(output_dir)
)

# 1. Create missing data objects if needed
if (!exists("pop_summary")) {
  pop_summary <- final_results %>%
    group_by(Pop) %>%
    summarise(
      mean_RMSE = mean(RMSE_MM_vs_CF),
      sd_RMSE = sd(RMSE_MM_vs_CF),
      n = n(),
      .groups = "drop"
    )
}

if (!exists("pairwise_with_meta")) {
  pairwise_with_meta <- final_results  # Use Chunk 3 output
}

# === 1. Summary Table ===
pop_table <- pop_summary %>%
  gt() %>%
  tab_header(title = "RMSE Summary by Population") %>%
  fmt_number(columns = where(is.numeric), decimals = 4) %>%
  cols_label(
    mean_RMSE = "Mean RMSE",
    sd_RMSE = "SD",
    n = "Sample Size"
  )

print(pop_table)

# === 2. RMSE vs Sample Size ===
p1 <- ggplot(pop_summary, aes(x = n, y = mean_RMSE)) +
  geom_point(aes(size = n), alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_RMSE-sd_RMSE, ymax = mean_RMSE+sd_RMSE), 
                width = 0.1, color = "gray50") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "steelblue") +
  labs(title = "Mean RMSE vs Population Sample Size", 
       subtitle = "Error bars show ±1 SD",
       x = "Sample Size (n)", 
       y = "Mean RMSE") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggsave(file.path(output_dir, "RMSE_vs_sample_size.png"), 
       plot = p1, width = 7, height = 5, dpi = 300)

# Check available columns in your data
glimpse(pairwise_with_meta)
# or
names(pairwise_with_meta)
# 3. Ridgeline Plot (using Pop_MM since populations should match within pairs)
p2 <- ggplot(pairwise_with_meta, aes(x = RMSE, y = Pop_MM, fill = Pop_MM)) +
  geom_density_ridges(
    alpha = 0.7, 
    scale = 0.9,
    quantile_lines = TRUE,
    quantiles = 2  # Shows median line
  ) +
  scale_fill_viridis_d(option = "plasma", name = "Population") +
  labs(
    title = "RMSE Distribution by Population",
    subtitle = "Using MM scan population classification",
    x = "RMSE (MM vs CF)", 
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggsave(file.path(output_dir, "RMSE_distribution_by_population.png"), 
       plot = p2, width = 7, height = 5, dpi = 300)

# === 4. ICC Plot (if data exists) ===
if (exists("icc_values")) {
  p3 <- ggplot(icc_values, aes(x = MM, y = CF)) +
    geom_hex(bins = 30) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_fill_viridis_c(option = "magma") +
    labs(title = "Shape Correspondence Between MM and CF",
         x = "MM Shape Vector", 
         y = "CF Shape Vector") +
    theme_minimal(base_size = 12)
  
  ggsave(file.path(output_dir, "MM_CF_correspondence.png"), 
         plot = p3, width = 6, height = 5, dpi = 300)
} else {
  warning("⚠️ icc_values not found - skipping correspondence plot")
}

# === 5. Displacement Statistics ===
if (exists("mean_disp")) {
  displacement_summary <- tibble(
    Metric = c("Min", "Mean", "Median", "SD", "Max", "95th Percentile"),
    Value = c(min(mean_disp), mean(mean_disp), median(mean_disp),
              sd(mean_disp), max(mean_disp), quantile(mean_disp, 0.95))
  )
  
  print(displacement_summary)
  write_csv(displacement_summary, file.path(output_dir, "displacement_summary.csv"))
} else {
  warning("⚠️ mean_disp not found - skipping displacement statistics")
}

cat("✅ Chunk 4 completed. Outputs saved to:", output_dir, "\n")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
